---
title: "The Beauty of R"
pagetitle: "Manasseh"
author: " "
output:
  html_document:
    fig_width: 5
    fig_height: 5
    fig_caption: true
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: TRUE
    theme: journal
    highlight: tango
---

<link rel="stylesheet" href="styles.css" type="text/css">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# recode variables

- Happy Coding !!!

In most survey projects, you'll find the importance to recode variables. In this tutorial, I will explain the steps needed to recode a variable from number to characters and vice versa.

I’ll create a data frame with 100 random sample with gender and marital status demographic variables.

## R-libraries used
```{r, warning = FALSE, message = FALSE, results = 'hide'}
library(dplyr)
library(ggplot2)
library(vtree)
```


## Generate data
```{r echo = TRUE, warning = FALSE, message = FALSE}
rm(list = ls())
set.seed(123)
gender = sample(0:1, 100, replace = TRUE)
marital.status = sample(1:3, 100, replace = TRUE)
dem.data = data.frame(gender, marital.status)
as_tibble(dem.data)
```

## Data Structure
```{r echo = TRUE, warning = FALSE, message = FALSE}
str(dem.data)
```

The structure of the data show both gender and marital status are numeric attributes, but the two are supposed to be factor variables. The as.factor () is useful in converting variables to factor.


```{r echo = TRUE, warning = FALSE, message = FALSE}
dem.data$gender = as.factor(dem.data$gender)
dem.data$marital.status = as.factor(dem.data$marital.status)
str(dem.data)
```

The gender and marital status are transformed from integer to factor variables. Gender has two levels (0 and 1) while marital status has three levels (1, 2, and 3).

## Visualization
```{r echo = TRUE, warning = FALSE, message = FALSE}
dem.data %>%
  ggplot(aes(gender)) +
  geom_bar(fill = "navyblue")

dem.data %>%
  ggplot(aes(marital.status)) +
  geom_bar(fill = "lawngreen")
```

As a data analyst you might have prior knowledge of the variable meanings as indicated in the metadata or codebook. However, a reader or the audience might not understand the two plot because they don't know what 0 and 1 mean in gender and 1, 2, and 3 in marital status, hence the need to recode variable.

## recode gender and marital status (number to factor)
```{r echo = TRUE, warning = FALSE, message = FALSE}
dem.data = dem.data %>%
  mutate(
    gender.cat = recode(gender, "0" = "male", "1" = "female")
  )

dem.data = dem.data %>%
  mutate(
    marital.status.cat = recode(marital.status, "1" = "single", "2" = "married", "3" = "divorced")
  )

dem.data %>%
  ggplot(aes(gender.cat)) +
  geom_bar(fill = "cornflowerblue") +
  xlab("Gender")

dem.data %>%
  ggplot(aes(marital.status.cat)) +
  geom_bar(fill = "brown") +
  xlab("Marital Status")
```


The audience can easily understand the two graphs as the labels are defined.

## recode gender and marital status (factor to number)
```{r echo = TRUE, warning = FALSE, message = FALSE}
dem.data = dem.data %>%
  mutate(
    gender.num = recode(gender.cat, "male" = 0, "female" = 1)
  )

dem.data = dem.data %>%
  mutate(
    marital.status.num = recode(marital.status.cat, "single" = 1, "married" = 2, "divorced" = 3)
  )

dem.data %>%
  ggplot(aes(gender.num)) +
  geom_bar(fill = "purple4") +
  xlab("Gender")

dem.data %>%
  ggplot(aes(marital.status.num)) +
  geom_bar(fill = "goldenrod4") +
  xlab("Marital Status")

p = vtree(dem.data, "gender.cat marital.status.cat")
```


# impute missing data 

- using mean and median

Missing values or missing data can be defined as the data value which is not available for a specific variable (s) in the observation of interest. Missing data have significant effect on final interpretation of results obtained from the data with missing values. Specifically, missing data can reduces statistical power, cause bias in the estimation of parameters, and can reduce the representativeness of the samples. However, the problem can be sorted out using a number of imputation methods. The blog series will exploit mean and median missing value imputations methods.


I’ll create a data frame with 50 random sample with multivariate analysis test scores and Income variables with additional missing values.

## R-libraries used
```{r, warning = FALSE, message = FALSE, results = 'hide'}
library(dplyr)
```


## Generate data
```{r echo = TRUE, warning = FALSE, message = FALSE}
rm(list = ls())
set.seed(456)

n = 50
df = data.frame(
score = sample(c(40:100), n, rep = TRUE), 
income = sample(c(1200:5000), n, rep = TRUE)
)

df = apply (df, 2, function(x) {x[sample( c(1:n), floor(n/5))] <- NA; x} )
df = as.data.frame(df)
class(df)
str(df)
as_tibble(df)
colSums(is.na(df))
```

## Check columns with NAs
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Return the column names containing missing observations
list_na = colnames(df)[ apply(df, 2, anyNA) ]
list_na
```

## Create mean
```{r echo = TRUE, warning = FALSE, message = FALSE}
average_missing = apply(df[, colnames(df) %in% list_na],
      2,
      mean,
      na.rm =  TRUE)
average_missing
```

## Create median
```{r echo = TRUE, warning = FALSE, message = FALSE}
median_missing = apply(df[,colnames(df) %in% list_na],
      2,
      median,
      na.rm =  TRUE)
median_missing
```


## Replace NAs using mean
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Create a new variable with mean
df_replace_mean = df %>%
   mutate(replace_mean_score  = ifelse(is.na(score), average_missing[1], score),
   replace_mean_income = ifelse(is.na(income), average_missing[2], income))
tibble(df_replace_mean)
sum(is.na(df_replace_mean))
```


## Replacing NAs using median
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Create a new variable with median
df_replace_median = df %>%
            mutate(replace_median_score  = ifelse(is.na(score), median_missing[1], score), 
            replace_median_income = ifelse(is.na(income), median_missing[2], income))
tibble(df_replace_median)
sum(is.na(df_replace_median))
```

## direct NA impute using mean 
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Quick code to replace missing values with the mean
df_impute_mean = data.frame(
    sapply(
        df,
        function(x) ifelse(is.na(x),
            mean(x, na.rm = TRUE),
            x)))
tibble(df_impute_mean)
sum(is.na(df_impute_mean))
```

## direct NA impute using median
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Quick code to replace missing values with the median
df_impute_median = data.frame(
    sapply(
        df,
        function(x) ifelse(is.na(x),
            median(x, na.rm = TRUE),
            x)))
tibble(df_impute_median)
sum(is.na(df_impute_median))
```


## direct NA impute using mean for numeric attributes
```{r echo = TRUE, warning = FALSE, message = FALSE}
# Replace missing values in the numeric attributes using mean
df[sapply(df, is.numeric)] = lapply(df[sapply(df, is.numeric)], 
                                      function(x)
                                        ifelse(is.na(x), 
                                               mean(x, na.rm = TRUE), x))
tibble(df)
sum(is.na(df))
```

- Happy Coding !!!


## R-libraries used
```{r, warning = FALSE, message = FALSE, results = 'hide'}
library(reticulate)
library(dplyr)
```

## Generate data
```{r echo = TRUE, warning = FALSE, message = FALSE}
rm(list = ls())
set.seed(567)

n = 250
df = data.frame(
loan.amount = sample(c(800:1500), n, rep = TRUE)
)

# Add outliers in the data
df = df %>% 
  add_row(loan.amount = c(7000,7050))

as_tibble(df)
```


```{python, echo = FALSE}
print(r.df.shape)
print(r.df.info())
r.df.describe()
```

```{python, echo = FALSE}
import matplotlib.pyplot as plt
r.df.plot.hist(grid = False, bins = 20, rwidth = 0.9,
                   color = 'purple')
```



```{python, echo = TRUE}
# identify outliers with interquartile range
from numpy import percentile

# calculate interquartile range
q25, q75 = percentile(r.df.loc[:1], 25), percentile(r.df.loc[:1], 75)
iqr = q75 - q25
iqr

print('Percentiles: 25th=%.3f, 75th=%.3f, IQR=%.3f' % (q25, q75, iqr))

# calculate the outlier cutoff
cut_off = iqr * 1.5
lower, upper = q25 - cut_off, q75 + cut_off

lower
upper
```







